### 2017-11-08 paper/sections/1_introduction.tex
paper: cleanup intro, pull in evaluation/related work, re-org evaluation
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

In this paper we present a detailed analysis of how the ParSplice application
accesses key-value pairs over the course of a long running simulation across a
variety of initial conditions. We reason that limiting the size of a cache on a
single node saves memory and sacrifices negligible performance. This type of
analysis (1) shows the capacity and resource requirements of a single node and
(2) will help inform our load balancing policies for when we switch to a
distributed key-value store back-end to store EOM minima. We need to know when
and how to partition the keyspace: a smaller cache hurts performance because
key-value pairs need to be retrieved from other nodes while a larger cache has
higher memory pressure.

@@

%First, we implement the ``how much" abstraction as an LRU and then we focus on
%the ``when" abstraction. 
%Manageable: abstracts away complexities of the system (pass around to others,
%use different strategies)

>>>>

\begin{figure}[t]
  \noindent\includegraphics[width=0.5\textwidth]{figures/motivation-regimes.png}\\
  \caption{The keyspace activity for ParSplice using two different growth
rates.  The line shows the rate that EOM minima values are retrieved from the
key-value store (\(y1\) axis) and the points along the bottom show the number
of keys accessed in a 1 second sliding window (\(y2\) axis).
\label{fig:motivation-regimes}}
\end{figure}

In this paper we present a detailed analysis of how the ParSplice application
accesses key-value pairs over the course of a long running simulation across a
variety of initial conditions.  This type of analysis (1) shows the capacity
and resource requirements of a single node and (2) will help inform our load
balancing policies for when we switch to a distributed key-value store back-end
to store EOM minima. We need to know when and how to partition the keyspace: a
smaller cache hurts performance because key-value pairs need to be retrieved
from other nodes while a larger cache has higher memory pressure.

For example, Figure~\ref{fig:motivation-regimes} shows the
cache activity of a ParSplice node and demonstrates that small changes to the
rates at which new atoms enter the simulation (\(\Delta\)) can have a strong
effect on the timing and frequency with which new EOM minima are discovered and
referenced.  The ParSplice caches on each node are trimmed when memory pressure
reaches a threshold but the number of unique keys accessed (black points
along bottom of both graphs) in Figure~\ref{fig:motivation-regimes} suggests
that a small cache of EOM minima should be sufficient.

@@

%Manageable: abstracts away complexities of the system (pass around to others,
%use different strategies)

### 2017-11-08 paper/sections/4_dynamic-policies.tex
paper: cleanup intro, pull in evaluation/related work, re-org evaluation
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

\section{Mantle Brains: Tools We Plug In to Detect ``when"}
\label{sec:mantle-brains}

Requirements: run online, be fast enough to run as often as we want to detect
regimes

\subsection{System Specific Knowledge}

{\it e.g.}, request rate, unique keys in a sliding window, bandwidth
capabilties. For example, we know that LevelDB cannot handle high IO request
rates.

\subsubsection{Request Rate}

\subsubsection{Belady's Min}

\subsection{Domain Specific Knowledge}

{\it e.g.}, ParSplice key access locality.

\subsubsection{Regime Detection}

@@

\begin{figure}[tb]
\begin{minted}[linenos]{lua}

-- assume that (ts, keyid) are in a table
local function when()

  if servers[whoami]["load"] > target then
    if servers[whoami]["load"] > absorb_target then
      WRstate(1)
    end
    if RDstate() == 1 then
      return true
    end
  end
  return false
end
\end{minted}
\caption{ParSplice cache management policy.\label{lua:parsplice}}
\end{figure}

\subsubsection{Trajectory Length}

\subsection{Failed, Overcomplicated Brains}
These techniques proliferated more, less transparent knobs
\begin{itemize}
  \item Statistics
  \item Calculus
  \item K-Means
  \item DBScan
  \item Anomaly Detection
\end{itemize}

@@

How Much: cache policy from past, regime detection

>>>>

\section{Cache Management Using System Architecture Knowledge}
\label{sec:cache-management-using-system-archiecture-knowledge}

Using the Mantle policy engine, we test a variety of cache management tools and
algorithms on a single in-memory database node using the keyspace analysis in
Section~\S\ref{sec:parsplice-keyspace-analysis}. These strategies are
implemented as the ``when" and ``how much" policies from the Mantle API; the
``where" callback does not make sense for a single node (segments are either in
the cache or they are not). The evaluation metric is the accuracy and runtime
of each strategy; the strategy should be accurate enough so as to sacrifice
negligible performance and fast tenough to run as often as we want to detect
regimes.

First we sized the cache according to our system specific knowledge. By
``system", we mean the hardware and software of the storage hierarchy. We look
at request rate, unique keys in a sliding window, and bandwidth capabilties. For
example, we know that LevelDB cannot handle high IO request rates.

% technical details
In the original ParSplice implementation, each cache node uses an unlimited
amount of memory to store segment coordinates. We limit the size of the cache
using an LRU eviction policy, where the penalty for a cache miss is retrieving
the data from the persistent database.  We evict keys (if necessary) at every
operation instead of when segments complete because the cache fills up too
quickly otherwise.

% results: cache size trade-offs
\begin{figure}[t]
  \noindent\includegraphics[width=0.5\textwidth]{figures/methodology-tradeoff.png}\\

  \caption{The performance and resource utilization trade-off for different
cache sizes. ``Baseline" is
ParSplice unmodified and the ``Policy: Constrain Cache Size" graph limits the
size of the cache to save memory.  \label{fig:methodology-tradeoff}}

\end{figure}

The results for different cache sizes for a growth rate of \(\Delta_1\) over a
2.5 hour run across 256 workers is shown in
Figure~\ref{fig:methodology-tradeoff}.  ``Baseline" is the performance of
unmodified ParSplice  measured in trajectory duration (\(y\) axis) and
utilization is measured with memory footprint of just the cache (\(y2\) axis).
The other graph shares the \(y\) axis and shows the trade-off of constraining
the cache to different sizes.  The error bars are the standard deviation of 3
runs.

% results: raw numbers
Although the keyspace grows to 150K, a 100K key cache achieves 99\% of the
performance. Decreasing the cache degrades performance and predictability.
While this result is not unexpected, it nonetheless achieves our goal of
showing the benefits of load balancing keys across nodes and that smaller
caches on each node are an effective way to save memory without completely
sacrificing performance.

% why is the performance lower for smaller caches?
Despite the memory savings, our results suggest that dynamic load balancing
policies could save even more memory.  Figure~\ref{fig:methodology-tradeoff}
show that a 100K key cache is sufficient as a static policy but the top graph
in Figure~\ref{fig:motivation-regimes} indicates that the cache size could be
much smaller. That graph shows that the beginning of the run is characterized
by many reads to a small set of keys and the end sees much lower reads per
second to a larger keyspace. Specifically, it shows only about 100 keys as
active in the latter half of the run.

After analyzing traces, we see that the 100 key cache is insufficient because
the persistent database cannot service the read-write traffic.  According to
Figure~\ref{fig:motivation-regimes}, the read requests arrive at 750 reads per
second in addition to the writes that land in each tier (about 300 puts/second,
some redundant). This traffic triggers a LevelDB compaction and reads block,
resulting in very slow progress.  Traces verify this hypothesis and show reads
getting backed up as the read/write ratio increases. To recap, small caches
incur too much load on the persistent database  at the beginning of the run but
should suffice after the initial read flash crowd passes because
the keyspace is far less active. This suggests a two-part load balancing
policy.

% what is mantle
To explore dynamic load balancing policies ({\it i.e.} policies that change
during the run), we use the Mantle approach.  Mantle is a framework built on the
Ceph file system that lets administrators control file system metadata load
balancing policies. The basic premise is that load balancing policies can be
expressed with a simple API consisting of ``when", ``where", and ``how much".
The succinctness of the API lets users inject multiple, dynamic policies.

% Why is this a good idea
Although ParSplice does not use a distributed file system, its workload is very
similar because the minima key-value store responds to small and frequent
requests, which results in hot spots and flash crowds.  Modern distributed file
systems have found efficient ways to measure, migrate, and partition metadata
load and have shown large performance gains and
better scalability~\cite{zheng:pdsw2014-batchfs, zheng:pdsw2015-deltafs,
grider:pdsw2015-marfs, ren:sc2014-indexfs, patil:fast2011-giga+,
brandt:msst2003-lh}.  Previous work quantified the speedups achieved with
Mantle and formalized balancers that were good for file systems.

\begin{figure}[t]
  \noindent\includegraphics[width=0.5\textwidth]{figures/methodology-tradeoff-dynamic.png}\\
  \caption{The performance and resource utilization trade-off of using a
  dynamic load balancing policy that switches to a constrained cache policy after absorbing
  the initial burstiness of the workload. The sizes of these smaller caches are
  on the \(x\) axis.  \label{fig:methodology-tradeoff-dynamic}}
\end{figure}

Figure~\ref{fig:methodology-tradeoff-dynamic} shows the results of using the
Mantle API to program a dynamic load balancing policy into ParSplice that
switches between two policies:

\begin{itemize}
  \item unlimited growth policy: cache increases on every write
  \item \(n\) key limit policy: cache constrained to \(n\) keys
\end{itemize}

We trigger the policy switch at 100K keys to absorb the flash crowd at the
beginning of the run. Once triggered, keys are evicted to bring the size of the
cache down to the threshold.  In the bar chart, the cache sizes for the \(n\)
key limit policy are along the \(x\) axis.

% results: same level of performance can be achieved 
The dynamic policies show better performance than the single \(n\) key
policies. The performance and memory utilization for a 100K key cache size is
the same as the 100K bar in Figure~\ref{fig:methodology-tradeoff-dynamic} but
the rest reduce the size of the keyspace after the read flash crowd.  We see
the worst performance when the engine switches to the 10 key limit policy,
which achieves 94\% of the performance while only using 40KB of memory.

% caveats: it is calculating 90% of the trajectory, memory value reported is final
\subsubsection*{Caveats}

The results in Figure~\ref{fig:methodology-tradeoff-dynamic} are slightly
deceiving for three reason: (1) segments take longer to generate later in the
run, (2) the memory footprint is the value at the end of 2.5 hours, and (3)
this policy only works well for the 2.5 hour run.  For (1), the curving down of
the simulation vs. wall-clock time is shown in
Figure~\ref{fig:methodology-trajectory}; as the nanoparticle grows it takes
longer to generate segments so by the time we reach 2 hours, over 90\% of the
trajectory is already generated.  For (2), the memory footprint is around 0.4GB
until we reach 100K key switch threshold. In
Figures~\ref{fig:methodology-tradeoff}
and~\ref{fig:methodology-tradeoff-dynamic} we plot the final value. For (3),
Figure~\ref{fig:methodology-trajectory} shows that the cache fills up with 100K
keys at time 7200 seconds and its size is reduced to the size listed in the
legend.  The curves stay close to ``Unlimited" for up to an hour after the
cache is reduced but eventually flatten out as the persistent database gets
overloaded. 10K and 100K follow the ``Unlimited" curve the longest and are
sufficient policies for the 2.5 hour runs but anything longer would need a
different dynamic load balancing policy.

\begin{figure}[t]
  \noindent\includegraphics[height=5cm,width=0.4\textwidth]{figures/methodology-trajectory.png}\\
  \caption{The rate that the trajectory is computed decays over time (which is
  expected) but this skews the performance improvements in
  Figure~\ref{fig:methodology-tradeoff-dynamic}. Our dynamic policy works for 2.5
  hour jobs but not for 4 hour jobs.  \label{fig:methodology-trajectory}}
\end{figure}

% but the result is still valid
Despite these caveats, the result is still valid: we found a dynamic load
balancing policy that absorbs the cost of a high read throughput on a small
keyspace and reduces the memory pressure for a 2.5 hour run. Our experiments
show the effectiveness of the load balancing policy engine we integrated into
ParSplice, not that we were able to identify the best policy for all system
setups ({\it i.e.} different ParSplice parameters, number of worker tasks, and
job lengths).  To solve that problem, we need a way to identify what thresholds
we should use for different job permutations.

\section{Cache Management Using Domain-Specific Application Knowledge}
\label{sec:cache-management-using-domain-specific-knowledge}
\begin{figure}[t]
  \noindent\includegraphics[height=5cm,width=0.4\textwidth]{figures/keyspace-zoomed.png}\\

  \caption{Key activity for a 4 hour run shows groups of accesses to the same
  subset of keys. Detecting this ``access regimes" leads to a more accurate cache
  management strategy.\label{fig:keyspace-zoomed}}

\end{figure}

Feeding domain-specific knowledge about the ParSplice application into a policy
leads to more a accurate cache management strategy.
Figure~\ref{fig:keyspace-zoomed} shows which keys (\(y\) axis) are accessed by
the ParSplice tasks over time (\(x\) axis). The groups of accesses to a subset
of the keys, or ``access regimes", occur because molecules are stuck in deep
trajectories. Recall that the in-memory database stores the molecules' EOM
minima, which is the smallest effective energy that a molecule observes during
its trajectory. So molecules stuck in deep trajectories explore the same minima
until they can escape to a new set of states. This exploration of the same set
of states is called a superbasin.

Detecting these superbasins can lead to more effective cache management
strategies because the height of the keyspace accesses is ``how much" of cache
to evict and the width of the keyspace accesses is ``when" to evict values from
the cache.  The zoomed portion of Figure~\ref{fig:keyspace-zoomed} shows how a
single superbasin affects the key accesses. Moving along the \(x\) axis shows
that the number of unique keys accessed over time grows while moving along the
\(y\) axis shows that early keys are accessed more often.  Overall, superbasins
are never re-visited because the simulation only adds molecules; we can never
reach a state with less molecules. This is why keys are never re-accessed.
Despite these patterns, the following characteristics of superbasins make it
hard to detect them:

\begin{itemize}

  \item superbasin key accesses are random and there is no threshold ``minimum distance
  between key access" that indicates we have moved on to a new superbasin

  \item superbasins change immediately

  \item the number of keys a superbasin accesses differs from other superbasins

\end{itemize}

Below we describe the policies we implemented and deployed using Mantle.

\subsection{Failed Strategies}
These techniques proliferated more knobs that obfuscated the problem. 
\begin{itemize}
  \item Statistics
  \item Calculus
  \item K-Means
  \item DBScan
  \item Anomaly Detection
\end{itemize}

\subsection{Regime Detection}

@@

\subsection{Trajectory Length}

@@

How Much: cache policy from past, regime detection, Belady's Min

### 2017-11-08 paper/sections/6_conclusion.tex
paper: cleanup intro, pull in evaluation/related work, re-org evaluation
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

@@

\section{Conclusion}
\begin{figure}[tb]
  \noindent\includegraphics[width=0.5\textwidth]{figures/keyspace-analysis_size.png}\\
  \caption{\label{fig:keyspace-analysis_size}}
\end{figure}
\begin{figure}[tb]
  \noindent\includegraphics[width=0.5\textwidth]{figures/keyspace-analysis_locality.png}\\
  \caption{\label{fig:keyspace-analysis_locality}}
\end{figure}
\begin{figure}[tb]
  \noindent\includegraphics[width=0.5\textwidth]{figures/keyspace-analysis_throughput.png}\\
  \caption{\label{fig:keyspace-analysis_throughput}}
\end{figure}
\begin{figure}[tb]
  \noindent\includegraphics[width=0.5\textwidth]{figures/keyspace-analysis_cachesize.png}\\
  \caption{\label{fig:keyspace-analysis_cachesize}}
\end{figure}
\begin{figure}[tb]
  \noindent\includegraphics[width=0.5\textwidth]{figures/keyspace-analysis_cachesize-dynamic.png}\\
  \caption{\label{fig:keyspace-analysis_cachesize-dynamic}}
\end{figure}
\begin{figure}[tb]
  \noindent\includegraphics[width=0.5\textwidth]{figures/keyspace-analysis_cachesize-caveats.png}\\
  \caption{\label{fig:keyspace-analysis_cachesize-caveats}}
\end{figure}

>>>>

\section{Related Work}

Key-value storage organizations for scientific applications is a field gaining
rapid interest. In particular, the analysis of the ParSplice keyspace and the
development of an appropriate scheme for load balancing is a direct response to
a case study for computation caching in scientific
applications~\cite{jenkins:ipdsw17-mochi}. In that work the authors motivated
the need for a flexible load balancing \emph{microservice} to efficiently scale
a memoization microservice. Our work is also heavily influenced by the
Malacology project~\cite{sevilla:eurosys17-malacology} which seeks to provide
fundamental services from within the storage system ({\it e.g.}, consensus) to
the application.

State-of-the-art distributed file systems partition write-heavy workloads and
replicate read-heavy workloads, similar to the approach we are advocating
here.  IndexFS~\cite{ren:sc2014-indexfs} partitions directories and clients
write to different partitions by grabbing leases and caching ancestor metadata
for path traversal. ShardFS takes the replication approach to the extreme by
copying all directory state to all nodes. The Ceph file system
(CephFS)~\cite{weil:sc2004-dyn-metadata, weil:osdi2006-ceph} employs both
techniques to a lesser extent; directories can be replicated or sharded but the
caching and replication policies are controlled with tunable parameters.  These
systems still need to be tuned by hand with {\it ad-hoc} policies designed for
specific applications.  Setting policies for migrations is arguably more
difficult than adding the migration mechanisms themselves.  For example,
IndexFS/CephFS use the GIGA+~\cite{patil:fast2011-giga} technique for
partitioning directories at a \emph{predefined} threshold. Mantle makes headway
in this space by providing a framework for exploring these policies, but does
not attempt anything more sophisticated (e.g., machine learning) to create
these policies.

% ml and autotuning
Auto-tuning is a well-known technique used in
HPC~\cite{behzad:sc2013-autotuning, behzad:techreport2014-io-autotuning}, big
data systems systems~\cite{herodotou_starfish_2011}, and
databases~\cite{schnaitter_index_2009}.  Like our work, these systems focus on
the physical design of the storage ({\it e.g.} cache size) but since we focused
on a relatively small set of parameters (cache size, migration thresholds), we
did not need anything as sophisticated as the genetic algorithm used
in~\cite{behzad:sc2013-autotuning}.  We cannot drop these techniques into
ParSplice because the magnitude and speed of the workload hotspots/flash crowds
makes existing approaches less applicable.

Our plan is to use MDHIM~\cite{greenberg:hotstorage2015-mdhim} as our back-end
key-value store because it was designed for HPC and has the proper mechanisms
for migration already implemented.

@@

\section{Conclusion}
